---
title: "PS270 Final Project"
author: Olivia Palmberg
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction 

In this project I have aimed to explore how education has affected voting patterns. The question I worked to answer is: As education increases, are individuals more or less in support of the democratic party. To uncover this, I looked at different demographics and how they affected voting results in the 2022 Democratic House race. My main hypothesis was that as education increases, support for the Democratic Party will increase. 

This research question is important because demographics play a very prominent role in the way that individuals vote and affiliate with different political parties. This question works to uncover a core driver of political behavior: how individual's backgrounds shape the way that they vote. Looking specifically at education helps explain larger patterns in elections in the way that education can affect how people access and process information. Furthermore, the way we access information has a large impact on our beliefs and party/candidate preferences as well. 


## Data Section 

To measure the support for the Democratic party among different education levels, I will be taking the data and creating four different conditions. I will categorize each individual into "HS or less", "College, under 4 years", "Bachelor's degree", and "Postgrad". Education is the explanatory variable, as each category/level will yield a different outcome of percentage in support for the 2022 Democratic House. The outcome variable is the percentage votes per each education level, reflecting support for the 2022 Democratic House. 

An outcome of higher percentages for each education increase would prove my hypothesis to be correct, while an outcome showing variability or decrease of percentages for each education increase would prove my hypothesis to be incorrect. 

```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(broom)
```

```{r}
vote_data <-
  haven::read_sav("ATP W117.sav") |>
  mutate(
    education = case_when(
      F_EDUCCAT2 %in% c(1, 2) ~ "HS or less",
      F_EDUCCAT2 %in% c(3, 4) ~ "College, under 4 years",
      F_EDUCCAT2 %in% c(5) ~ "Bachelor's Degree",
      F_EDUCCAT2 %in% c(6) ~ "Postgrad",
      .default = NA
    ) |>
      fct_relevel(c("HS or less", "College, under 4 years")),
    congvote_dem = if_else(CONGPOST_W117 == 2, TRUE, FALSE)
  ) |>
  select(id = QKEY, education, congvote_dem) |>
  filter(!is.na(congvote_dem),
         !is.na(education))
vote_data
```

Here I did some data cleaning to create table that shows individuals under the different education levels, and if they voted in support of the 2022 Democratic House or not. If they voted in support of the Democratic Party they are labeled "TRUE", if they did not then they are labeled "FALSE". 

```{r}
edu_vote <- vote_data |>
  group_by(education) |>
  summarize(percentage_dem = mean(congvote_dem) * 100) 
knitr::kable(edu_vote, digits = 2)
```

Here I created a clean table showing the average percentage of support for the 2022 Democratic House for each education level. 


## Results Section

```{r}
twentytwo_dem_house <- vote_data |>
  group_by(education) |>
  summarize(dem_pct = 100 * mean(congvote_dem, na.rm = TRUE)) |>
ggplot(mapping = aes(x = education, y =  dem_pct)) +
  geom_col() +
  labs(x = "Education Levels", y = "Percentage Votes for 2022 Democratic House", title = "Distribution of 2022 Democratic House Votes Accross Education Levels")
twentytwo_dem_house
```
Here I plotted the distribution of the percentage that each education level voted democrat in the House Election. The distribution shows that as education gets higher, it seems that there is generally stronger support for the democratic party because the percentage of votes is increasing. However, it seems that college under four years has a slightly lower support for the democratic party than high school or less. 


```{r}
fit <- lm(congvote_dem ~ education, data = vote_data) 
fit

tidy(fit) |>
  knitr::kable(digits = 3)
```
Here I made a regression that showed education levels and support for the Democratic party. This uses high school or less as the baseline intercept and it seems that support goes down by about .06 at college under 4 years. Then from there the support increases for both Bachelor's degree and Postgrad. 


## 

```
