---
title: "PS270 Final Project"
author: Olivia Palmberg
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to recent elections and the how the voting trends have changed. I would specifically like to look into how different factors effect voting behaviors. An example of this would be how education plays a role in voting. 

How has education affected voting trends and patterns in recent elections? Based on data from recent elections, it seems that as education gets higher, people tend to vote for the Democratic Party as opposed to the Republican Party. To measure this, we will be analyzing different levels of education in the categories of "HS or less", "Some College", "4 year College", and "Postgrad". Education has our explanatory variables. Our outcome variable is the percentage towards that party/candidate. We will be looking at how education changes the official results in terms of percentage for Democratic and Republican presidential candidates from past years. The observed pattern from the data points to higher education changing percentages to be higher for Democratic candidates. 

## 2022 Democratic House Vote 

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
vote_data <-
  haven::read_sav("ATP W117.sav") |>
  mutate(
    education = case_when(
      F_EDUCCAT2 %in% c(1, 2) ~ "HS or less",
      F_EDUCCAT2 %in% c(3, 4) ~ "College, under 4 years",
      F_EDUCCAT2 %in% c(5) ~ "Bachelor's Degree",
      F_EDUCCAT2 %in% c(6) ~ "Postgrad",
      .default = NA
    ) |>
      fct_relevel(c("HS or less", "College, under 4 years")),
    congvote_dem = if_else(CONGPOST_W117 == 2, TRUE, FALSE)
  ) |>
  select(id = QKEY, education, congvote_dem) |>
  filter(!is.na(congvote_dem),
         !is.na(education))
vote_data
```

Here I did some data cleaning to create a clea table that shows different education levels and if they voted for the democratic party or not. 

## Distribution of 2022 Democratic House Votes Accross Education Levels 

```{r}
twentytwo_dem_house <- vote_data |>
  group_by(education) |>
  summarize(dem_pct = 100 * mean(congvote_dem, na.rm = TRUE)) |>
ggplot(mapping = aes(x = education, y =  dem_pct)) +
  geom_col() +
  labs(x = "Education Levels", y = "Percentage Votes for 2022 Democratic House", title = "Distribution of 2022 Democratic House Votes Accross Education Levels")
print(twentytwo_dem_house)
```
Here I plotted the distribution of the percentage that each education level voted democrat in the House Election. The distribution shows that as education gets higher, it seems that there is generally stronger support for the democratic party because the percentage of votes is increasing. However, it seems that college under four years has a slightly lower support for the democratic party than highnschool or less. 

## Regression 

```{r}
lm(congvote_dem ~ education, data = vote_data) |> summary()
```
Here I made a regression that showed education levels and support for the Democratic party. This uses high school or less as the baseline intercept and it seems that support goes down by about .06 at colege under 4 years. Then from there the support increases for both Bachelor's degree and Postgrad. 


```
